Appendix: A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-03 appendix.tex
rearranged main paper and appendix into separate files again

\articletype{ARTICLE TEMPLATE}

\title{A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics
with the Lineup Protocol}

@@

\begin{abstract}
Abstract to fill.
\end{abstract}

\begin{keywords}
statistical graphics; data visualization; visual inference; hypothesis
testing; reression analysis; cognitive perception; simulation; practical
significance; effect size
\end{keywords}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\begin{quote}
\emph{``Since all models are wrong the scientist must be alert to what
is importantly wrong.''} \citep{box1976science}
\end{quote}

Diagnostics are the key to determining whether there is anything
importantly wrong with a model. In linear regression analysis, residuals
from the model fit are commonly used. Residuals summarise what is not
captured by the model, and thus provide the capacity to identify what
might be wrong.

We can assess residuals in multiple ways. Residuals may be plotted as a
histogram or quantile-quantile plot to examine the distribution. Using
the classical normal linear regression model as an example, if the
distribution is symmetric and unimodal, we consider it to be well
behaved. However, if the distribution is skewed, bimodal, multimodal, or
contains outliers, there is a cause for concern. One could also inspect
the distribution by conducting a goodness of fit test, such as the
Shapiro-Wilk Normality test \citep{shapiro1965analysis}.

More typically, residuals will be plotted, as a scatter plot against the
predicted values and each of the explanatory variables to scrutinize
their relationships. If there are any visually discoverable patterns,
the model is potentially misspecified. In general, one looks for
noticeable departures from the model like non-linear dependency or
heteroskedasticity. However, correctly judging a residual plot where no
pattern exists can be a painstakingly difficult task for humans. It is
especially common, particularly among students, to misinterpret patterns
that are random noise and random deviation from a model
\citep{loy2021bringing}. It is also possible to conduct hypothesis tests
for non-linear dependence \citep{ramsey_tests_1969}, and use a
Breusch-Pagan test \citep{breusch_simple_1979} for heteroskedasticity.

Abundance of literature describe appropriate diagnostic methods for
linear regression, e.g. \citet{draper1998applied},
\citet{montgomery1982introduction}, \citet{belsley_regression_1980},
\citet{cook_applied_1999} and \citet{cook1982residuals}. All these
writings consider plotting residuals to be a standard technique that
should be examined routinely in all regression modelling problems. In
addition, \citet{draper1998applied} and \citet{belsley_regression_1980}
believe that residual plots are usually revealing when the assumptions
are violated. \citet{cook_applied_1999} thinks formal tests and
graphical procedures are complementary and both have a place in residual
analysis, but they focus on graphical methods rather than on formal
testing, as they are easier to use. \citet{montgomery1982introduction}
even suggests that residual plots are more informative in most practical
situations than the corresponding formal tests, and statistical tests on
regression model residuals are not widely used based on their
experience.

A common guidance by experts is that optimal method for diagnosing model
fits is by plotting the data. The persistence of this advice to check
the plots is curious, and investigating why this might be common advice
is the subject of this paper. The paper is structured as follows. The
next section describes the background on the types of departures that
one expects to detect, and outlines a formal statistical process for
reading residual plots, called visual inference. Section
\ref{experimental-design} details the experimental design to compare the
decisions made by formal hypothesis testing, and how humans would read
diagnostic plots. The results are reported in Section \ref{results}. We
conclude with a discussion of future work, in particular, how the
responsibility for residual plot reading might be relieved.

\hypertarget{background}{%
\section{Background}\label{background}}

\hypertarget{departures-from-good-residual-plots}{%
\subsection{Departures from good residual
plots}\label{departures-from-good-residual-plots}}

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/residual-plot-common-departures-1}

@@

\caption{Example residual vs fitted value plots: (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted value plot.}\label{fig:residual-plot-common-departures}
\end{figure}

Graphical summaries in which residuals are plotted against fitted values
or other functions of the predictors that are approximately orthogonal
to the residuals are referred to as standard residual plots in
\citet{cook1982residuals}. Figure
\ref{fig:residual-plot-common-departures}A shows an ideal residual plot
where the residuals are evenly distributed at both sides of the
horizontal zero line, with no noticeable patterns.

There are various types of departures from an ideal residual plot.
Non-linearity, heteroskedasticity and non-normality are perhaps the
three mostly checked departures.

Non-linearity is a type of model misspecification caused by failing to
include higher order terms of the predictors in the regression equation.
Any non-linear functional form of residuals on fitted values in the
residual plot could be indicative of non-linearity. An example residual
plot containing visual pattern of non-linearity is shown in Figure
\ref{fig:residual-plot-common-departures}B. One can clearly observe the
``S-shape'' from the residual plot as the cubic term is not captured by
the misspecified model.

Heteroskedasticity refers to the presence of nonconstant error variance
in a regression model. It is mostly due to the strict but false
assumptions on the variance-covariance matrix of the error term. The
usual pattern of heteroskedasticity on a residual plot is the
inconsistent spread of the residuals across the horizontal axis.
Visually, it sometimes results in the so-called ``butterfly'' shape as
shown in Figure \ref{fig:residual-plot-common-departures}C, or the
``left-triangle'' and ``right-triangle'' shape where the smallest
variance occurs at one side of the horizontal axis.

Compared to non-linearity and heteroskedasticity, non-normality is
usually harder to detect from a residual plot since a scatter plot do
not readily reveal the marginal distribution. A favourable graphical
summary for this task is the quantile-quantile plot. As we mainly
discuss residual plots, non-normality will not be the focus of this
paper. For a consistent comparison, the residual plot of this departure
is still presented in Figure \ref{fig:residual-plot-common-departures}D.
When the number residuals below and above the horizontal axis are uneven
across the local regions along the \(x\)-axis, we expect that the
normality assumption is violated. For example, given a skewed error
distribution, there will be fewer data points and more outliers on one
side of the horizontal axis as shown in Figure
\ref{fig:residual-plot-common-departures}D.

\hypertarget{conventionally-testing-for-departures}{%
\subsection{Conventionally testing for
departures}\label{conventionally-testing-for-departures}}

Other than checking diagnostic plots, analysts may perform formal
hypothesis testing to detect model defects. A variety of tests can be
applied; each testing for a specific violation of the null hypothesis.
For example, the presence of heteroskedasticity can usually be tested by
applying the White test \citep{white_heteroskedasticity-consistent_1980}
or the Breusch-Pagan test \citep{breusch_simple_1979}, which are both
derived from the Lagrange multiplier test \citep{silvey1959lagrangian}
principle that relies on the asymptotic properties of the null
distribution. To test specific forms of non-linearity, one may apply the
F-test as a model structural test to examine the significance of
specific polynomial and non-linear forms of the predictors, or the
significance of proxy variables as in the Ramsey Regression Equation
Specification Error Test (RESET) \citep{ramsey_tests_1969}. The
Shapiro-Wilk test \citep{shapiro1965analysis} is the most widely used
test of non-normality included by many of the statistical software
programs. The Jarque--Bera test \citep{jarque1980efficient} is also used
to directly check whether the sample skewness and kurtosis match a
normal distribution.

We apply the RESET test, Breusch-Pagan test and Shapiro-Wilk test to the
residual plots shown in Figure
\ref{fig:residual-plot-common-departures}; the results of this is shown
in Table \ref{tab:example-residual-plot-table}. The Breusch-Pagan test
and the Shapiro-Wilk test both reject \(H_0\) for residual plots that
display non-linearity and heteroscedasticity, even though these are not
the original intention of the test. As discussed in
\citet{cook1982residuals}, most residual-based tests for a particular
type of departure from model assumptions are also sensitive to other
types of departures. It is likely \(H_0\) is correctly rejected but for
the wrong reason, a phenomenon known as the ``Type III error''.
Additionally, outliers will often incorrectly trigger the rejection of
\(H_0\) despite when majority of the residuals are well-behaved
\citep{cook_applied_1999}. Furthermore, with a sufficiently large sample
size, residual-based tests may reject \(H_0\) due to a slight departure
that is of little practical significance. These can be largely avoided
in diagnostic plots as experienced analysts can evaluate the
acceptability of assumptions flexibly, even in the presence of outliers
and slight departures.

@@

\caption{\label{tab:example-residual-plot-table}Statistical significance testing for departures from good residuals for plots in Figure \ref{fig:residual-plot-common-departures}. Shown are the $p$-values calculated for the RESET, the Breusch-Pagan and the Shapiro–Wilk tests. The good residual plot (A) is judged a good residual plot, as expected, by all tests. The non-linearity (B) is detected by all tests, as might be expected given the extreme structure.}
\centering
\begin{tabular}[t]{llrrr}
\toprule
Plot & Departures & RESET & Breusch-Pagan & Shapiro–Wilk\\
\midrule
A & None & 0.779 & 0.133 & 0.728\\
B & Non-linearity & \em{0.000} & \em{0.000} & \em{0.039}\\
C & Heteroskedasticity & 0.658 & \em{0.000} & \em{0.000}\\
D & Non-normality & 0.863 & 0.736 & \em{0.000}\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{visual-test-procedure-based-on-lineups}{%
\subsection{Visual test procedure based on
lineups}\label{visual-test-procedure-based-on-lineups}}

One may argue that reading diagnostic plots is to some extent subjective
and indecisive compared to those rigorous statistical procedures as it
relies on graphical perception - human ability to interpret and decode
the information embedded in graph \citep{cleveland_graphical_1984}.
Further, the degree of the presence of the visual features typically can
not be measured quantitatively and objectively, which may lead to over
or under-interpretations of the data. For instance, people
over-interpret the separation between gene groups in a two-dimensional
projection from a linear discriminant analysis when in fact there are no
differences in the expression levels between the gene groups and
separation is not an uncommon occurrence
\citep{roy_chowdhury_using_2015}.

Visual inference was first introduced in a 1999 Joint Statistical
Meetings (JSM) talk with the title ``Inference for Data Visualization''
by \citet{buja_inference_1999} as an idea to address the issue of valid
inference for visual discoveries of data plots. Later,
\citet{buja_statistical_2009} proposed the lineup protocol as a visual
test inspired by the ``police lineup'' or ``identity parade'' which is
the act of asking the eyewitness to identify criminal suspect from a
group of irrelevant people. The protocol consists of \(m\) randomly
placed plots, where one plot is the data plot, and the remaining
\(m - 1\) null plots have the identical graphical procedure except the
data has been replaced with data consistent with \(H_0\), that the model
is correctly specified. Then, an observer who have not seen the data
plot will be asked to point out the most different plot from the lineup.
Under \(H_0\), it is expected that the data plot would have no
distinguishable difference from the null plots, and the probability that
the observer correctly picks the data plot is \(1/m\). If one rejects
\(H_0\) as the observer correctly picks the data plot, then the Type I
error of this test is \(1/m\).

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/first-example-lineup-1}

@@

\caption{Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data (plot $2^2 + 2$, exhibiting non-linearity) is embedded among 19 null plots, where the residuals are simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting plot $2^2 + 2$.}\label{fig:first-example-lineup}
\end{figure}

Figure \ref{fig:first-example-lineup} is an example of a lineup
protocol. If the data plot at position \(2^2 + 2\) is identifiable, then
it is evidence for the rejection of \(H_0\). In fact, the actual
residual plot is obtained from a misspecified regression model with
missing non-linear terms.

Data used in the \(m - 1\) null plots needs to be simulated. In
regression diagnostics, sampling data consistent with \(H_0\) is
equivalent to sampling data from the assumed model. As
\citet{buja_statistical_2009} suggested, \(H_0\) is usually a composite
hypothesis controlled by nuisance parameters. Since regression models
can have various forms, there is no general solution to this problem,
but it sometimes can be reduced to a so called ``reference
distribution'' by applying one of the three methods: (i) sampling from a
conditional distribution given a minimal sufficient statistic under
\(H_0\), (ii) parametric bootstrap sampling with nuisance parameters
estimated under \(H_0\), and (iii) Bayesian posterior predictive
sampling. The conditional distribution given a minimal sufficient
statistic is the best justified reference distribution among the three
\citep{buja_statistical_2009}. Essentially, null residuals can be
simulated by regressing \(N\) i.i.d standard normal random draws on the
predictors, then rescaling it by the ratio of residual sum of square in
two regressions.

The effectiveness of lineup protocol for regression analysis is
validated by \citet{majumder_validation_2013} under relatively simple
settings with up to two predictors. Their results suggest that visual
tests are capable of testing the significance of a single predictor with
a similar power as a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a conventional test,
and they do not expect the visual test to perform equally well as the
conventional test. In their third experiment, where there is not a
conventional test, visual test outperforms the conventional test for a
large margin. This is encouraging, as it promotes the use of visual
inference in situations where there are no existing statistical testing
procedures. Visual inference have also been integrated into diagnostic
of hierarchical linear models by \citet{loy2013diagnostic},
\citet{loy2014hlmdiag} and \citet{loy2015you}. They use lineup protocols
to judge the assumption of linearity, normality and constant error
variance for both the level-1 and level-2 residuals.

\hypertarget{calculation-of-statistical-significance-and-test-power}{%
\section{Calculation of statistical significance and test
power}\label{calculation-of-statistical-significance-and-test-power}}

\hypertarget{what-is-being-tested}{%
\subsection{What is being tested?}\label{what-is-being-tested}}

In diagnosing a model fit from residuals, we are generally interested in
\(H_0:\) \emph{The regression model is correctly specified} against the
broad alternative \(H_a:\) \emph{The regression model is misspecified}.

However, it is practically impossible to test this specific \(H_0\) with
conventional tests, which are constructed to measure specific
departures. For example, the RESET test is formulated as
\(H_0:\gamma_1 = \gamma_2 = \gamma_3 = 0\) against
\(H_a: \gamma_1 \neq 0 \text{ or } \gamma_2 \neq 0 \text{ or } \gamma_3 \neq 0\),
from
\(y = \tau_0 + \sum_{i=1}^{p}\tau_px_p +\gamma_1\hat{y}^2 + \gamma_2\hat{y}^3 + \gamma_3\hat{y}^4 + u, ~~u \sim N(0, \sigma_u^2)\).
Similarly, the Breusch-Pagan test is designed to specifically test
\(H_0:\) \emph{error variances are all equal}
(\(\zeta_i=0 \text{ for } i=1,..,p\)) versus the alternative \(H_a:\)
\emph{that the error variances are a multiplicative function of one or
more variables} (\(\text{at least one } \zeta_i\neq 0\)) from
\(e^2 = \zeta_0 + \sum_{i=1}^{p}\zeta_i x_i + u, ~ u\sim N(0,\sigma_u^2)\).

One of the potential benefits of the visual test, based on the lineup
protocol, is that it should be able to detect a range of departures from
good residuals.

\hypertarget{statistical-significance}{%
\subsection{Statistical significance}\label{statistical-significance}}

In hypothesis testing, a \(p\)-value is defined as the probability of
observing test results as least as extreme as the observed result given
\(H_0\) is true. Conventional hypothesis tests usually have an existing
method to derive or compute \(p\)-value based on the null distribution.
What we need to discuss in the following is the method to estimate
\(p\)-value for a visual test.

Within the context of visual inference, by involving \(k\) independent
observers, the visual \(p\)-value can be interpreted as the probability
of having as many or more subjects detect the data plot than the
observed result.

Let \(X_j = \{0,1\}\) be a Bernoulli random variable denoting whether
subject \(j\) correctly detecting the data plot, and
\(X = \sum_{j=1}^{K}X_j\) be the number of observers correctly picking
the data plot. Then, by imposing a relatively strong assumption on the
visual test that all \(K\) evaluations are fully independent, under
\(H_0\), \(X \sim \mathrm{Binom}_{K,1/m}\). Therefore, the \(p\)-value
of a lineup of size \(m\) evaluated by \(K\) observer is given as
\(P(X \geq x) = 1 - F(x) + f(x)\), where \(F(.)\) is the binomial
cumulative distribution function, \(f(.)\) is the binomial probability
mass function and \(x\) is the realization of number of observers
correctly picking the data plot \citep{majumder_validation_2013}.

As pointed out by \citet{vanderplas2021statistical}, this basic binomial
model does not take into account the possible dependencies in the visual
test due to repeated evaluations of the same lineup. And it is
inapplicable to visual test where subjects are asked to select one or
more ``most different'' plots from the lineup.
\citet{vanderplas2021statistical} summarises three common scenarios in
visual inference: (1) \(K\) different lineups are shown to \(K\)
subjects, (2) \(K\) lineups with different null plots but the same data
plot are shown to \(K\) subjects, and (3) the same lineup is shown to
\(K\) subjects. Out of these three scenarios, Scenario 3 is the most
common in previous studies as it puts the least constraints on the
experiment design. For Scenario 3, \citet{vanderplas2021statistical}
models the probability of a plot \(i\) being selected from a lineup as
\(\theta_i\), where \(\theta_i \sim Dirichlet(\alpha)\) for
\(i=1,...,m\) and \(\alpha > 0\). The number of times plot \(i\) being
selected in \(K\) evaluations is denoted as \(c_i\). In case subject
\(j\) makes multiple selections, \(1/s_j\) will be added to \(c_i\)
instead of one, where \(s_j\) is the number of plots subject \(j\)
selected for \(j=1,...K\). This ensures \(\sum_{i}c_i=K\). Since we are
only interested in the selections of the data plot \(i\), the marginal
model can be simplified to a beta-binomial model and thus the visual
\(p\)-value is given as

\begin{equation} \label{eq:pvalue-beta-binomial}
P(C \geq c_i) = \sum_{x=c_i}^{K}{K \choose x}\frac{B(x + \alpha, K - x + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+
\end{equation}

\noindent where \(B(.)\) is the beta function defined as

\begin{equation} \label{eq:betafunction}
B(a, b) = \int_{0}^{1}t^{\alpha - 1}(1-t)^{b-1}dt,\quad \text{where}\quad a,b>0.
\end{equation}

Note that Equation \ref{eq:pvalue-beta-binomial} given in
\citet{vanderplas2021statistical} only works with non-negative integer
\(c_i\). We extend the equation to non-negative real number \(c_i\) by
applying a linear approximation

\begin{equation} \label{eq:pvalue-beta-binomial-approx}
P(C \geq c_i) = P(C \geq \lceil c_i \rceil) + (\lceil c_i \rceil - c_i) P(C = \lfloor c_i \rfloor), \quad \text{for}\quad c_i \in \mathbb{R}_0^+,
\end{equation}

where \(P(C \geq \lceil c_i \rceil)\) is calculated using Equation
\ref{eq:pvalue-beta-binomial} and \(P(C = \lfloor c_i \rfloor)\) is
calculated by

\begin{equation} \label{eq:pmf-beta-binomial}
P(C = c_i) = {K \choose c_i}\frac{B(c_i + \alpha, K - c_i + (m - 1)\alpha)}{B(\alpha, (m-1)\alpha)},\quad \text{for} \quad c_i \in \mathbb{Z}_0^+.
\end{equation}

Besides, the parameter \(\alpha\) used in Equation
\ref{eq:pvalue-beta-binomial} and \ref{eq:pmf-beta-binomial} is usually
unknown and hence needs to be estimated from the survey data. For low
values of \(\alpha\), only a few plots are attractive to the observers
and tend to be selected. For higher values of \(\alpha\), the
distribution of the probability of each plot being selected is more
even. \citet{vanderplas2021statistical} defines that a plot is
\(c\)-interesting if \(c\) or more participants select the plot as the
most different. Given the definition, The expected number of plots
selected at least \(c\) times, \(E[Z_c]\), is calculated as

\begin{equation} \label{eq:c-interesting-expectation}
E[Z_c(\alpha)] = \frac{m}{B(\alpha, (m-1)\alpha)}\sum_{\lceil c \rceil}^{K}{K \choose x} B(x + \alpha, K - x + (m-1)\alpha).\end{equation}

With Equation \ref{eq:c-interesting-expectation}, \(\alpha\) can be
estimated using maximum likelihood estimation. But for precise estimate
of \(\alpha\), additional human responses to Rorschach lineups, which is
a type of lineup that consists of plots constructed from the same null
data generating mechanism, are required.

\hypertarget{power-of-the-tests}{%
\subsection{Power of the tests}\label{power-of-the-tests}}

The power of a model misspecification test is the probability that
\(H_0\) is rejected given the regression model is misspecified in a
specific way. It is an important indicator when one is concerned about
whether model assumptions have been violated. Although in practice, one
might be more interested in knowing how much the residuals deviate from
the model assumptions, and whether this deviation is of practical
significance.

The power of a conventional hypothesis test is affected by both the true
parameter \(\boldsymbol{\theta}\) and the sample size \(n\). These two
can be quantified in terms of effect size \(E\) to measure the strength
of the residual departures from the model assumptions. Details about the
effect size is provided in Section \ref{effect-size} after the
introduction of the simulation model used in our human subject
experiment. The theoretical power of a test is sometimes not a trival
solution, but it can be estimated if the data generating process is
known. We use a predefined model to generate a large set of simulated
data under different effect sizes, and record if the conventional test
rejects \(H_0\). The probability of the conventional test rejects
\(H_0\) is then fitted by a logistic regression formulated as

\begin{equation} \label{eq:logistic-regression-1-1}
Pr(\text{reject}~H_0|H_1,E) = \Lambda\left(log\left(\frac{0.05}{0.95}\right) + \beta_1 E\right),
\end{equation}

\noindent where \(\Lambda(.)\) is the standard logistic function given
as \(\Lambda(z) = exp(z)/(1+exp(z))\). The effect size \(E\) is the only
predictor and the intercept is fixed to \(log(0.05/0.95)\) so that
\(\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05\), which is the desired
significance level.

The power of a visual test on the other hand, may additionally depend on
the ability of the particular subject, as the skill of the individual
may affect the number of observers who identify the data plot from the
lineup \citep{majumder_validation_2013}. To address this issue,
\citet{majumder_validation_2013} models the probability of a subject
\(j\) correctly picking the data plot from a lineup \(l\) using a
mixed-effect logistic regression, with subjects treated as random
effect. Then, the estimated power of a visual test evaluated by a single
subject is the predicted value obtained from the mix-effect model.
However, this mix-effect model does not work with scenario where
subjects are asked to select one or more most different plots. In this
scenario, having the probability of a subject \(j\) correctly picking
the data plot from a lineup \(l\) is insufficient to determine the power
of a visual test because it does not provide information about the
number of selections made by the subject for the calculation of the
\(p\)-value (See Equation \ref{eq:pvalue-beta-binomial-approx}).
Therefore, we directly estimate the probability of a lineup being
rejected by assuming that individual skill has negligible effect on the
variation of the power. This assumption is not necessary true, but it
helps simplifying the model structure, thereby obviate a costly
large-scale experiment to estimate complex covariance matrices. The same
model given in Equation \ref{eq:logistic-regression-1-1} is applied to
model the power of a visual test.

To study various factors contributing to the power of both tests, the
same logistic regression model is fit on different subsets of the
collated data grouped by levels of factors. These include the
distribution of the fitted values, type of the simulation model and the
shape of the residual departures.

\hypertarget{experimental-design}{%
\section{Experimental design}\label{experimental-design}}

An experiment is conducted in three data collection periods to
investigate the difference between conventional hypothesis testing and
visual inference in the application of linear regression diagnostics.
Two types of departures, non-linearity and heteroskedasticity, are
collected during data collection periods I and II. The data collection
period III was designed primarily to measure human responses to null
lineups so that the parameter \(\alpha\) in Equation
\ref{eq:pvalue-beta-binomial} can be estimated. Additional lineups for
both non-linearity and heteroskedasticity, using uniform fitted value
distribution, were included so that the participants were evaluating
some lineups with signal also. It would be too frustrating for
participants to only be assigned lineups with all null plots. Overall,
we collected 7974 evaluations on 1152 unique lineups performed by 443
subjects throughout three data collection periods.

\hypertarget{simulating-departures-from-good-residuals}{%
\subsection{Simulating departures from good
residuals}\label{simulating-departures-from-good-residuals}}

\hypertarget{non-linearity}{%
\subsubsection{Non-linearity}\label{non-linearity}}

Data collection period I is designed to study the ability of human
subjects to detect the effect of a non-linear term \(\boldsymbol{z}\)
constructed using Hermite polynomials on random vector
\(\boldsymbol{x}\) formulated as

\begin{align} \label{eq:nonlinearity-model}
\boldsymbol{y} &= 1 + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1), \\
\boldsymbol{z} &= g(\boldsymbol{z}_{raw}, 1), \\
\boldsymbol{z}_{raw} &= He_j(g(\boldsymbol{x}, 2)),
\end{align}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\), \(\boldsymbol{x}_{raw}\),
\(\boldsymbol{z}_{raw}\) are vectors of size \(n\), \(He_{j}(.)\) is the
\(j\)th-order probabilist's Hermite polynomials,
\(\varepsilon \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)\), and
\(g(\boldsymbol{x}, k)\) is a scaling function to enforce the support of
the random vector to be \([-k, k]^n\) defined as

\begin{equation} \label{eq:scaling-function}
g(\boldsymbol{x}, k) = (\boldsymbol{x} - min(\boldsymbol{x}))/max(\boldsymbol{x} - min(\boldsymbol{x}))2k - k, \quad \text{for} \quad k > 0. 
\end{equation}

According to \citet{abramowitz1964handbook}, Hermite polynomials were
initially defined by \citet{de1820theorie}, but named after Hermite
\citep{hermite1864nouveau} because of the unrecognisable form of
Laplace's work. When simulating \(\boldsymbol{z}_{raw}\), function
\texttt{hermite} from the R package \texttt{mpoly} \citep{mpoly} is used
to generate Hermite polynomials.

The null regression model used to fit the realizations generated by the
above model is formulated as

\begin{equation} \label{eq:null-model}
\boldsymbol{y} = \beta_0 + \beta_1 \boldsymbol{x} + \boldsymbol{u},
\end{equation}

\noindent where
\(\boldsymbol{u} \sim N(\boldsymbol{0}, \sigma^2\boldsymbol{I}_n)\).

Since \(z = O(x^j)\), for \(j > 1\), \(z\) is a higher order term leaves
out by the null regression, which will lead to model misspecification.

Visual patterns of non-linearity are simulated using four different
orders of probabilist's Hermite polynomials (\(j = 2, 3, 6, 18\)). (A
summary of the factors is given in Table \ref{tab:model-factor-table}.)
The values of \(j\) is chosen so that distinct shapes of non-linearity
are included in the residual plot. These include ``U'', ``S'', ``M'' and
``triple-U'' shape as shown in Figure
\ref{fig:different-shape-of-herimite}. A greater value of \(j\) will
result in a curve with more turning points. It is expected that the
``U'' shape will be the easiest one to detect because complex shape
tends to be concealed by cluster of data points.

Figure \ref{fig:example-poly-lineup} demonstrates one of the lineups
used in non-linearity detection. This lineup is produced by the
non-linearity model with \(j = 6\). The data plot location is
\(2^3 - 4\). All five subjects correctly identify the data plot from
this lineup.

\caption{\label{tab:model-factor-table}Levels of the factors used in data collection periods I, II, III.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{rr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rcrr|rr|rc}
\toprule
\multicolumn{2}{c}{Non-linearity} & \multicolumn{2}{c}{Heteroskedasticity} & \multicolumn{2}{c}{Common} \\
\cmidrule(l{3pt}r{3pt}){1-2} \cmidrule(l{3pt}r{3pt}){3-4} \cmidrule(l{3pt}r{3pt}){5-6}
\multicolumn{1}{c}{Poly Order ($j$)} & \multicolumn{1}{c}{SD ($\sigma$)} & \multicolumn{1}{c}{Shape ($a$)} & \multicolumn{1}{c}{Ratio ($b$)} & \multicolumn{1}{c}{Size ($n$)} & \multicolumn{1}{c}{Distribution of fitted values} \\
\cmidrule(l{3pt}r{3pt}){1-1} \cmidrule(l{3pt}r{3pt}){2-2} \cmidrule(l{3pt}r{3pt}){3-3} \cmidrule(l{3pt}r{3pt}){4-4} \cmidrule(l{3pt}r{3pt}){5-5} \cmidrule(l{3pt}r{3pt}){6-6}
2 & 0.25 & -1 & 0.25 & 50 & Uniform\\
3 & 1.00 & 0 & 1.00 & 100 & Normal\\
6 & 2.00 & 1 & 4.00 & 300 & Skewed\\
18 & 4.00 &  & 16.00 &  & Discrete uniform\\
 &  &  & 64.00 &  & \\
\bottomrule
\end{tabular}}
\end{table}

@@

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-shape-of-herimite-1}

@@

\caption{Polynomial forms generated for the residual plots used to assess detecting non-linearity. The four shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.}\label{fig:different-shape-of-herimite}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-sigma-1}

\caption{Examining the effect of $\sigma$ on the signal strength in the non-linearity detection, for $n=300$, uniform fitted value distribution and the "U" shape. As $\sigma$ increases the signal strength decreases, to the point that the "U" is almost unrecognisable when $\sigma=4$.}\label{fig:different-sigma}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/example-poly-lineup-1}

\caption{One of the lineups containing non-linearity patterns used in data collection period I. Can you spot the most different plot? The data plot is positioned at $2^3 + 1$.}\label{fig:example-poly-lineup}
\end{figure}

\hypertarget{heteroskedasticity}{%
\subsubsection{Heteroskedasticity}\label{heteroskedasticity}}

Data collection period II is designed to study the ability of human
subjects to detect the appearance of a heteroskedasticity pattern under
a simple linear regression model setting:

\begin{align} \label{eq:heter-model}
\boldsymbol{y} &= 1 + \boldsymbol{x} + \boldsymbol{\varepsilon},\\
\boldsymbol{x} &= g(\boldsymbol{x}_{raw}, 1),\\
\boldsymbol{\varepsilon} &\sim N(\boldsymbol{0}, 1 + (2 - |a|)(\boldsymbol{x} - a)^2b \boldsymbol{I}), 
\end{align}

\noindent where \(\boldsymbol{y}\), \(\boldsymbol{x}\),
\(\boldsymbol{\varepsilon}\) are vectors of size \(n\) and \(g(.)\) is
the scaling function defined in Equation \ref{eq:scaling-function}.

The null regression model used to fit the realizations generated by the
above model is formulated exactly the same as Equation
\ref{eq:null-model}.

For \(b \neq 0\), the variance-covariance matrix of the error term
\(\boldsymbol{\varepsilon}\) is correlated with the predictor
\(\boldsymbol{x}\), which will lead to the presence of
heteroskedasticity. Visual patterns of heteroskedasticity are simulated
using three different shapes (\(a\) = -1, 0, 1). (A summary of the
factors can be found in Table \ref{tab:model-factor-table}.)

Since \(supp(X) = [-1, 1]\), choosing \(a\) to be \(-1\), \(0\) and
\(1\) can generate ``left-triangle'', ``butterfly'' and
``right-triangle'' shape as displayed in Figure
\ref{fig:different-shape-of-heter}. The term \((2 - |a|)\) maintains the
magnitude of residuals across different values of \(a\).

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-shape-of-heter-1}

\caption{Heteroskedasticity forms used in the experiment. Three different shapes ($a = -1, 0, 1$) are used in the experiment to create left-triangle, "butterfly" and "right-triangle" shapes, respectively.}\label{fig:different-shape-of-heter}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/different-b-1}

\caption{Five different values of $b$ are used in heteroskedasticity simulation to control the strength of the signal. Larger values of $b$ yield a bigger difference in variation, and stus stronger heteroskedasticity signal.}\label{fig:different-b}
\end{figure}

{\centering \includegraphics[width=1\linewidth]{paper_files/figure-latex/example-heter-lineup-1}

\caption{One of the lineups containing heteroskedasticity pattern used in data collection period II. Can you spot the most different plot? The data plot is positioned at $3^3 - 3^2$}\label{fig:example-heter-lineup}
\end{figure}

An example lineup of this model used in data collection period II is
shown in Figure \ref{fig:example-heter-lineup} with \(a = -1\). The data
plot location is \(2^4 + 2\). Nine out of 11 subjects correctly identify
the data plot from this lineup.

\hypertarget{factors-common-to-both-data-collection-periods}{%
\subsubsection{Factors common to both data collection
periods}\label{factors-common-to-both-data-collection-periods}}

>>>>

\articletype{}

\title{Appendix: A Plot is Worth a Thousand Tests: Assessing Residual
Diagnostics with the Lineup Protocol}

@@

\hypertarget{additional-details-of-testing-procedures}{%
\section{Additional details of testing
procedures}\label{additional-details-of-testing-procedures}}

\hypertarget{effect-size-derivation}{%
\subsection{Effect size derivation}\label{effect-size-derivation}}

\hypertarget{sensitivity-analysis-for-alpha}{%
\subsection{\texorpdfstring{Sensitivity analysis for
\(\alpha\)}{Sensitivity analysis for \textbackslash alpha}}\label{sensitivity-analysis-for-alpha}}

@@

{\centering \includegraphics[width=1\linewidth]{appendix_files/figure-latex/sensitivity-1}

@@

@@

\caption{\label{tab:sensitivity}Examining how decisions might change if $\hat{\alpha}$ was different. Percentage of lineups that where the $p$-value would switch to above or below 0.05, when $\hat{\alpha}$ is varied by a fractional amount.}
\centering
\begin{tabular}[t]{rll}
\toprule
fraction & reject to not reject & not reject to reject\\
\midrule
0.500 & 2.51 & 0\\
0.750 & 1.43 & 0\\
0.875 & 1.08 & 0\\
1.125 & 0 & 1.08\\
1.250 & 0 & 1.43\\
\addlinespace
1.500 & 0 & 1.79\\
\bottomrule
\end{tabular}
\end{table}

\hypertarget{effect-of-number-of-evaluations-on-the-power-of-a-visual-test}{%
\subsection{Effect of number of evaluations on the power of a visual
test}\label{effect-of-number-of-evaluations-on-the-power-of-a-visual-test}}

@@

{\centering \includegraphics[width=0.7\linewidth]{appendix_files/figure-latex/n_eval-1}

@@

\hypertarget{power-of-a-reset-test-under-different-auxiliary-regression-formulas}{%
\subsection{Power of a RESET test under different auxiliary regression
formulas}\label{power-of-a-reset-test-under-different-auxiliary-regression-formulas}}

@@

{\centering \includegraphics[width=1\linewidth]{appendix_files/figure-latex/reset-1}

@@

\newpage
